Open-source tools in R for landscape ecology

### 2021-01-06 paper/paper.Rmd
some edits

---
title: Open-source tools in R for landscape ecology
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
# thanks: |
#    Grants or other notes about the article that should go on the front
#    page should be placed here. General acknowledgments should be placed at the
#    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes M. Signer
  address: Department of Wildlife Sciences, University of Göttingen, Büsgenweg 3, 37077 Göttingen, Germany
  email: j.m.signer@gmail.com
- name: Laura J. Graham
  address: Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- R packages
- reproducibility
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

While *R* has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* and *PROJ* libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
<!--JN: references??-->
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.

@@

Creating maps is essential when working with spatial data.
It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
*R* allows to create two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g. changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function.
However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.

@@

Quick interactive visualization of spatial data can be done with the *mapview* package\cite{R-mapview}.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.

@@

MHKH and JN are authors of the *landscapemetrics* and *landscapetools* package.
JN is author of *belg* and *motif* package.
<!-- MH: I wasn't sure if we need to list this here, but it kind of felt like a not completely stupid idea. What do you think? -->

>>>>

---
title: Open-source tools in R for landscape ecology
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
# thanks: |
#    Grants or other notes about the article that should go on the front
#    page should be placed here. General acknowledgments should be placed at the
#    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- R packages
- reproducibility
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

While *R* has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL} and *PROJ* \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.

@@

Creating maps is essential when working with spatial data.
Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
*R* allows to create two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g. changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function.
However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.

@@

Quick interactive visualization of spatial data can be done with the *mapview* package\cite{R-mapview}.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
<!--- JS: I wonder if we should also cite js library? However, doing a quick google, I was not able to find a correct citation -->

@@

MHKH and JN are authors of the *landscapemetrics* and *landscapetools* package.
JN is author of *belg* and *motif* package. JS is author of the *amt* package. 
<!-- MH: I wasn't sure if we need to list this here, but it kind of felt like a not completely stupid idea. What do you think? -->
<!-- JS: I don't think its critical, but surely it doesn't do any arm listing this here -->

### 2021-01-06 paper/paper.tex
some edits

\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Johannes M. Signer \and  Laura J. Graham \and  }

\institute{
        Maximillian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
     \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Johannes M. Signer \at
     Department of Wildlife Sciences, University of Göttingen, Büsgenweg 3, 37077 Göttingen, Germany \\
     \email{\href{mailto:j.m.signer@gmail.com}{\nolinkurl{j.m.signer@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Laura J. Graham \at
     Geography, Earth \& Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom \\
     \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

While \emph{R} has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} and \emph{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, \emph{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine. Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.

@@

Creating maps is essential when working with spatial data.
It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
\emph{R} allows to create two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function.
However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.

@@

MHKH and JN are authors of the \emph{landscapemetrics} and \emph{landscapetools} package.
JN is author of \emph{belg} and \emph{motif} package.

>>>>

\author{  Maximillian H.K. Hesselbarth \and  Jakub Nowosad \and  Johannes Signer \and  Laura J. Graham \and  }

\institute{
        Maximillian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA \\
     \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Johannes Signer \at
     Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Büsgenweg 3, 37077 Göttingen, Germany \\
     \email{\href{mailto:jsigner@uni-goettingen.de}{\nolinkurl{jsigner@uni-goettingen.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Laura J. Graham \at
     Geography, Earth \& Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom \\
     \email{\href{mailto:l.graham@bham.ac.uk}{\nolinkurl{l.graham@bham.ac.uk}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

While \emph{R} has several built-in data structures, including vectors, matrices, data.frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \emph{R} packages (\cite{Lovelace2019}, page 10).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \emph{R} language.
Importantly, main \emph{R} packages for spatial data use the external \emph{GDAL} \cite{GDAL} and \emph{PROJ} \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, \emph{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine. Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.

@@

Creating maps is essential when working with spatial data.
Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
\emph{R} allows to create two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function.
However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.

@@

MHKH and JN are authors of the \emph{landscapemetrics} and \emph{landscapetools} package.
JN is author of \emph{belg} and \emph{motif} package. JS is author of the \emph{amt} package.
